---
title: "NHBE SARS-CoV-2 RNA-seq analysis"
author: "Meiyue Wang"
date: "2026-03-21"
output: html_document
---
# Project goal

This project analyzes RNA-seq data from GSE147507.
I focus on NHBE cells and compare SARS-CoV-2 infection with mock treatment.

# Load packages

```{r}
library(DESeq2)
library(tidyverse)
library(ggplot2)
```

# Read data

```{r}
counts <- read.table("data/raw/GSE147507_RawReadCounts_Human.tsv",
                      header = TRUE,
                      sep = "\t",
                      row.names = 1,
                      check.names = FALSE
                      )
dim(counts)
```
Data retrieved successfully 21797=gene number 78=sample number

# Select NHBE+Mock vs SARS-CoV-2

## Check column names
```{r}
colnames(counts)
```
## Select desired groups

```{r}
samples <- colnames(counts)

selected <- samples[
  grepl("Series1_NHBE", samples) &
  (grepl("Mock", samples) | grepl("SARS-CoV-2", samples))
]

selected
counts_sub <- counts[, selected]

dim(counts_sub)
colnames(counts_sub)
```
## Look at the dimensions 

```{r}
dim(counts_sub)
```

# Build sample metadata (样本信息表)

```{r}
coldata <- data.frame(
  row.names = colnames(counts_sub),
  condition = factor(
    c("control", "control", "control",
      "infected", "infected", "infected"),
    levels = c("control", "infected")
  )
)

coldata
```
# DESeq2 analysis

## Filter out low-expression genes

```{r}
keep <- rowSums(counts_sub >= 10) >= 3
counts_filtered <- counts_sub[keep, ]

dim(counts_filtered)
```
the number of rest genes = 12291

## DESeq2

```{r}
dds <- DESeqDataSetFromMatrix(
  countData = counts_filtered,
  colData = coldata,
  design = ~ condition
)

dds <- DESeq(dds)
res <- results(dds)

summary(res)
head(res)
```
上调基因：406 下调基因：200

# Visulization

```{r}
res_df <- as.data.frame(res)
res_df <- res_df[!is.na(res_df$padj), ]

res_df$significance <- "Not Sig"
res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange > 1] <- "Up"
res_df$significance[res_df$padj < 0.05 & res_df$log2FoldChange < -1] <- "Down"

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up" = "red", "Down" = "blue", "Not Sig" = "grey")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(
    title = "Volcano plot: SARS-CoV-2 vs Control (NHBE)",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  theme_minimal()
```
```














